---
title: "EDA Notebook"
author: "Shawal Fida"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    toc: true
    toc_depth: '2'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load Libraries
library(tidyverse)       # Data manipulation & visualization
library(skimr)           # Data summary
library(corrplot)        # Correlation heatmaps
library(naniar)          # Missing data visualization
library(janitor)         # Data cleaning

```


## Business Problem
Home Credit Group aims to broaden financial inclusion for unbanked populations by providing a positive and safe borrowing experience. The key challenge is accurately predicting which loan applicants will be able to repay their loans. This prediction is critical because:

1. Rejecting good applicants means lost business opportunities
2. Approving high-risk applicants may lead to financial losses and increased default rates.
3. Many potential customers lack conventional credit histories, making traditional scoring methods inadequate

The company currently uses various statistical and machine learning methods to make these predictions but believes there's room for improvement. Our goal, and therefore the purpose of this notebook, is to develop a model that can more accurately identify which clients are capable of repayment, allowing Home Credit to make better-informed lending decisions.

## Analytical Problem
Customer repayment probability will be generated using a supervised machine-learning model. The model will use as inputs data collected on past customers such as application, demographic, and historical credit behavior. The model will use this information to predict the probability that the customers will either repay or not repay their loans. 

## Description of the Data

The dataset used in this project is `application_train.csv`, which contains **loan application records** from Home Credit. The dataset includes **demographic, financial, and credit-related attributes** of applicants. 

Each row represents an **individual loan application**, with the `TARGET` variable indicating whether the applicant **defaulted on their loan** (`1`) or successfully repaid it (`0`).

### 📊 **Key Features in the Dataset**
The dataset consists of **308,000+ records and 122 columns**. Some of the most important features include:

| **Feature Name**        | **Description** |
|-------------------------|----------------|
| `SK_ID_CURR`           | Unique ID for each loan application. |
| `TARGET`               | Loan status (0 = Repaid, 1 = Defaulted). |
| `NAME_CONTRACT_TYPE`   | Type of loan (Cash loans or Revolving loans). |
| `CODE_GENDER`          | Gender of the applicant (M/F). |
| `FLAG_OWN_CAR`         | Whether the applicant owns a car (Y/N). |
| `FLAG_OWN_REALTY`      | Whether the applicant owns real estate (Y/N). |
| `CNT_CHILDREN`         | Number of children the applicant has. |
| `AMT_INCOME_TOTAL`     | Total annual income of the applicant. |
| `AMT_CREDIT`          | Total loan credit amount. |
| `AMT_ANNUITY`         | Annual loan repayment amount. |


## Download Data and inspect

```{r data loading, message = FALSE, warning = FALSE}

application_train = read_csv("application_train.csv", col_names = TRUE, show_col_types = FALSE)
application_test = read_csv("application_test.csv", col_names = TRUE, , show_col_types = FALSE)

```

## Dataset Structure and Summary

```{r structure, message = FALSE, warning = FALSE, results = 'hide'}

# Dataset structure
str(application_train) 

# Dataset summary
summary(application_train) 

```


## Convert TARGET variable to a factor

```{r factor target, message = FALSE, warning = FALSE}
# Create a working copy of the dataset
train <- application_train

# Convert TARGET variable to a factor
train$TARGET <- as.factor(train$TARGET)

# Verify the conversion
str(train$TARGET)
table(train$TARGET)  # Check class distribution
```


## Check Missing Values 

```{r missing values, message = FALSE, warning = FALSE}

# Function to calculate missing values
missing_values <- function(df) {
  mis_val <- colSums(is.na(df))
  mis_val_percent <- (mis_val / nrow(df)) * 100
  mis_table <- data.frame(Feature = names(mis_val),
                         Missing_Values = mis_val,
                         Percentage = mis_val_percent) %>%
    filter(Missing_Values > 0) %>%
    arrange(desc(Percentage))
  
  return(mis_table)
}

# Check missing values
train_missing <- missing_values(train)
test_missing <- missing_values(application_test)
head(train_missing, 10)  # Display top 10 features with missing values

# Plot missing values (Top 20)
missing_values(train) %>%
  head(20) %>%
  ggplot(aes(x = reorder(Feature, -Percentage), y = Percentage)) +
  geom_bar(stat = "identity", fill = "red") +
  coord_flip() +
  labs(title = "Top 20 Features with Missing Values",
       x = "Features",
       y = "Percentage of Missing Values")

```

# Dropping columns with more than 60% missing values

```{r drop values, message = FALSE, warning = FALSE}

# Define threshold
threshold <- 0.60  # 60% missing values threshold

# Calculate missing values percentage
missing_values_ratio <- colSums(is.na(train)) / nrow(train)
missing_values_test_ratio <- colSums(is.na(application_test)) / nrow(application_test)

# Select columns to keep (less than 60% missing values)
train <- train[, missing_values_ratio < threshold]
application_test <- application_test[, missing_values_test_ratio < threshold]

# Print remaining columns
cat("Number of remaining columns after dropping:", ncol(train), "\n")

```


# Handling remaining missing values

```{r impute values, message = FALSE, warning = FALSE}

# 1. Impute numeric columns with median
num_cols <- names(train)[sapply(train, is.numeric)]
num_cols_with_na <- names(which(sapply(train[num_cols], function(x) sum(is.na(x))) > 0))

for (col in num_cols_with_na) {
  med_val <- median(train[[col]], na.rm = TRUE)
  train[[col]][is.na(train[[col]])] <- med_val
}

# Impute for test dataset
num_cols_test <- names(application_test)[sapply(application_test, is.numeric)]
num_cols_test_with_na <- names(which(sapply(application_test[num_cols_test], function(x) sum(is.na(x))) > 0))
for (col in num_cols_test_with_na) {
  med_val <- median(application_test[[col]], na.rm = TRUE)
  application_test[[col]][is.na(application_test[[col]])] <- med_val
}

# 2. Impute categorical columns with most frequent value (mode)
cat_cols <- names(train)[sapply(train, is.character) | sapply(train, is.factor)]
cat_cols_with_na <- names(which(sapply(train[cat_cols], function(x) sum(is.na(x))) > 0))

for (col in cat_cols_with_na) {
  mode_val <- names(sort(table(train[[col]]), decreasing = TRUE))[1]
  train[[col]][is.na(train[[col]])] <- mode_val
}

# Impute for testing dataset
cat_cols_test <- names(application_test)[sapply(application_test, is.character) | sapply(application_test, is.factor)]
cat_cols_test_with_na <- names(which(sapply(application_test[cat_cols_test], function(x) sum(is.na(x))) > 0))

for (col in cat_cols_test_with_na) {
  mode_val <- names(sort(table(application_test[[col]]), decreasing = TRUE))[1]
  application_test[[col]][is.na(application_test[[col]])] <- mode_val
}

# Check if any missing values remain
any_missing_train <- any(is.na(train))
cat("Are there any missing values in train data?", any_missing_train, "\n")

```

## Outlier Analysis

```{r outliers, message = FALSE, warning = FALSE}

# Identify numeric columns (excluding ID and target variable)
num_cols <- names(train)[sapply(train, is.numeric)]
num_cols <- num_cols[!num_cols %in% c("SK_ID_CURR", "TARGET")]

# Function to detect outliers using IQR method
identify_outliers <- function(x) {
  q1 <- quantile(x, 0.25, na.rm = TRUE)
  q3 <- quantile(x, 0.75, na.rm = TRUE)
  iqr <- q3 - q1
  lower_bound <- q1 - 1.5 * iqr
  upper_bound <- q3 + 1.5 * iqr
  return(list(
    lower = lower_bound,
    upper = upper_bound,
    outliers = sum(x < lower_bound | x > upper_bound, na.rm = TRUE),
    outlier_prop = sum(x < lower_bound | x > upper_bound, na.rm = TRUE) / length(x[!is.na(x)])
  ))
}

# Calculate outliers for each numeric variable
outlier_summary <- data.frame(
  Feature = character(),
  Lower_Bound = numeric(),
  Upper_Bound = numeric(),
  Outlier_Count = numeric(),
  Outlier_Proportion = numeric(),
  stringsAsFactors = FALSE
)

for (col in num_cols) {
  if (length(unique(train[[col]])) > 1) { # Skip constant columns
    outlier_info <- identify_outliers(train[[col]])
    outlier_summary <- rbind(outlier_summary, data.frame(
      Feature = col,
      Lower_Bound = outlier_info$lower,
      Upper_Bound = outlier_info$upper,
      Outlier_Count = outlier_info$outliers,
      Outlier_Proportion = outlier_info$outlier_prop,
      stringsAsFactors = FALSE
    ))
  }
}

# Sort by proportion of outliers
outlier_summary <- outlier_summary %>% arrange(desc(Outlier_Proportion))

# Display top variables with outliers
head(outlier_summary, 10)

# Visualize distributions of top variables with outliers
for (i in 1:min(5, nrow(outlier_summary))) {
  col_name <- outlier_summary$Feature[i]
  p <- ggplot(train, aes(x = .data[[col_name]])) +
    geom_histogram(fill = "blue", alpha = 0.7, bins = 30) +
    geom_vline(xintercept = outlier_summary$Lower_Bound[i], color = "red", linetype = "dashed") +
    geom_vline(xintercept = outlier_summary$Upper_Bound[i], color = "red", linetype = "dashed") +
    labs(title = paste("Distribution of", col_name),
         subtitle = "Red lines show outlier thresholds",
         x = col_name,
         y = "Count")
  print(p)
}
```

## Outliers handling

```{r outlier handling, message = FALSE, warning = FALSE}
# Strategy 1: Capping (Winsorization) for key financial variables
key_financial_vars <- c("AMT_INCOME_TOTAL", "AMT_CREDIT", "AMT_ANNUITY", "AMT_GOODS_PRICE")

for (col in key_financial_vars) {
  if (col %in% names(train)) {
    # Calculate outlier bounds
    q1 <- quantile(train[[col]], 0.25, na.rm = TRUE)
    q3 <- quantile(train[[col]], 0.75, na.rm = TRUE)
    iqr <- q3 - q1
    lower_bound <- q1 - 1.5 * iqr
    upper_bound <- q3 + 1.5 * iqr
    
    # Count values outside bounds
    n_lower <- sum(train[[col]] < lower_bound, na.rm = TRUE)
    n_upper <- sum(train[[col]] > upper_bound, na.rm = TRUE)
    
    # Apply capping
    train[[col]] <- ifelse(train[[col]] < lower_bound, lower_bound, train[[col]])
    train[[col]] <- ifelse(train[[col]] > upper_bound, upper_bound, train[[col]])
    
    cat("Capped", col, "- Values below lower bound:", n_lower, 
        "- Values above upper bound:", n_upper, "\n")
  }
}

## Do the same for testing dataset
for (col in key_financial_vars) {
  if (col %in% names(application_test)) {
    # Calculate outlier bounds
    q1 <- quantile(application_test[[col]], 0.25, na.rm = TRUE)
    q3 <- quantile(application_test[[col]], 0.75, na.rm = TRUE)
    iqr <- q3 - q1
    lower_bound <- q1 - 1.5 * iqr
    upper_bound <- q3 + 1.5 * iqr
    
    # Count values outside bounds
    n_lower <- sum(application_test[[col]] < lower_bound, na.rm = TRUE)
    n_upper <- sum(application_test[[col]] > upper_bound, na.rm = TRUE)
    
    # Apply capping
    application_test[[col]] <- ifelse(application_test[[col]] < lower_bound, lower_bound, train[[col]])
    application_test[[col]] <- ifelse(application_test[[col]] > upper_bound, upper_bound, train[[col]])
    
    cat("Capped", col, "- Values below lower bound:", n_lower, 
        "- Values above upper bound:", n_upper, "\n")
  }
}

# Strategy 2: Create binary flags for significant outliers in other variables
# Select variables with high outlier proportion (> 40%)
high_outlier_vars <- outlier_summary$Feature[outlier_summary$Outlier_Proportion > 0.4 & 
                                             !outlier_summary$Feature %in% key_financial_vars]

# Create flags for these variables
for (col in high_outlier_vars[1:10]) { # Limit to top 10 for efficiency
  if (col %in% names(train)) {
    # Get bounds from outlier summary
    lower <- outlier_summary$Lower_Bound[outlier_summary$Feature == col]
    upper <- outlier_summary$Upper_Bound[outlier_summary$Feature == col]
    
    # Create flag for outliers
    flag_name <- paste0(col, "_OUTLIER")
    train[[flag_name]] <- ifelse(train[[col]] < lower | train[[col]] > upper, 1, 0)
    
    cat("Created outlier flag for", col, "- Number of outliers:", sum(application_test[[flag_name]], na.rm = TRUE), "\n")
  }
}

# do same for testing dataset
for (col in high_outlier_vars[1:10]) { # Limit to top 10 for efficiency
  if (col %in% names(application_test)) {
    # Get bounds from outlier summary
    lower <- outlier_summary$Lower_Bound[outlier_summary$Feature == col]
    upper <- outlier_summary$Upper_Bound[outlier_summary$Feature == col]
    
    # Create flag for outliers
    flag_name <- paste0(col, "_OUTLIER")
    application_test[[flag_name]] <- ifelse(application_test[[col]] < lower | application_test[[col]] > upper, 1, 0)
    
    cat("Created outlier flag for", col, "- Number of outliers:", sum(application_test[[flag_name]], na.rm = TRUE), "\n")
  }
}


# Check distributions after handling outliers for a key variable
cat("\nVisualize distribution after handling outliers for", key_financial_vars[1], "\n")
p <- ggplot(train, aes(x = .data[[key_financial_vars[1]]])) +
  geom_histogram(fill = "green", alpha = 0.7, bins = 30) +
  labs(title = paste("Distribution of", key_financial_vars[1], "After Capping"),
       x = key_financial_vars[1],
       y = "Count")
print(p)

# Compare original vs. flagged distribution for a high-outlier variable
if (length(high_outlier_vars) > 0) {
  outlier_var <- high_outlier_vars[1]
  flag_var <- paste0(outlier_var, "_OUTLIER")
  
  cat("\nRelationship between", outlier_var, "outliers and target variable\n")
  table_result <- table(train[[flag_var]], train$TARGET)
  print(table_result)
  
  # Calculate proportion of defaults by outlier status
  prop_table <- prop.table(table_result, margin = 1)
  cat("\nProportion of defaults by outlier status:\n")
  print(prop_table[,2])
}

```




